
# Elevation data of Experiment plot in North Agronomy Farm

The goal of this activity is to visualize the elevation data of an experimental site in the North Agronomy Farm at Kansas State University. Also my goal is to predict elevation with different models and infer at what location we observe the lowest and highest elevation point.

## Librarires

```{r message=FALSE, warning=FALSE}
library(sf)
library(sp)
library(raster)
library(tidyverse)
```

## Kansas boundary

```{r message=FALSE, warning=FALSE}
# Download shapefile of Kansas from census.gov
download.file("http://www2.census.gov/geo/tiger/GENZ2015/shp/cb_2015_us_state_20m.zip", destfile = "states.zip")
unzip("states.zip")
sf.us <- st_read("cb_2015_us_state_20m.shp")
sf.kansas <- sf.us[48,6]
sf.kansas <- as(sf.kansas, 'Spatial')
plot(sf.kansas,main="",col="white")

```

## My data: North Agronomy Farm

```{r fig.keep=2:4, message=FALSE, warning=FALSE}
# Make shapefile of study area around Manhattan KS
url <- "https://www.dropbox.com/scl/fi/b7d9254b8f26mhu6pc2km/Experiment_walk.gpx?rlkey=nmph96rzfz8rxmwxsxg3wn4p1&dl=1"
pt.study.area <- st_read(dsn=url,layer="track_points")
sf.study.area  <- st_polygon(list(rbind(st_coordinates(pt.study.area),st_coordinates(pt.study.area)[1,])))
sf.study.area <- st_buffer(sf.study.area, .00006)
sf.study.area <- st_sf(st_sfc(sf.study.area), crs = crs(sf.kansas))

# Kansas + North farm data
{plot(sf.kansas,main="",col="white",xlim=c(-102.0517,-94.59193),ylim=c(36.99308,40.00308))
plot(sf.study.area, add=TRUE, col="red")}

```

## Extract elevation

```{r message=FALSE, warning=FALSE}
url <- "https://www.dropbox.com/scl/fi/b7d9254b8f26mhu6pc2km/Experiment_walk.gpx?rlkey=nmph96rzfz8rxmwxsxg3wn4p1&dl=1"

pt.elev <- st_read(dsn=url,layer="track_points")
pt.elev <- pt.elev[,4] # Keep only elevation
pt.elev <- pt.elev[-c(1:15),]
#pt.elev <- rbind(pt.elev,pt.study.area[,4])
```

## Visualize boundary + elevation

```{r}
{plot(sf.study.area)
plot(pt.elev,add=TRUE)}
hist(pt.elev$ele,col="grey")
summary(pt.elev$ele)

ggplot() +
  geom_sf(data=sf.study.area) +
  labs(subtitle = "North Agronomy Farm")+
  geom_sf(data=pt.elev, aes(color = ele), size = 2)+
  scale_color_gradient(low="blue", high="red", name = "Elevation (m)")+
  theme(panel.background = element_blank(),
        axis.text = element_blank())
# Visualize with satelite image 
mapview::mapview(sf.study.area)  

```

```{r}
# Transform pointfiles and shapefiles to utm zone to a planar coordinate reference system
pt.elev.utm <- st_transform(pt.elev,CRS("+proj=utm +zone=14 +datum=WGS84  +units=m"))
sf.study.area.utm <- st_transform(sf.study.area,CRS("+proj=utm +zone=14 +datum=WGS84  +units=m"))
```

```{r}
# Make data.frame for statistical analysis
df.elev <- data.frame (elev = pt.elev$ele,
                       long = st_coordinates(pt.elev)[,1],
                       lat = st_coordinates(pt.elev)[,2],
                       s1 = st_coordinates(pt.elev.utm)[,1],
                       s2 = st_coordinates(pt.elev.utm)[,2])
```

## Models

### 1. Linear with iid errors

```{r}
# Statistical analysis 1: non-hierarchical linear model with iid errors
m1 <- lm(elev~s1+I(s1^2)+s2+I(s2^2),data=df.elev)
# Make raster of study area to be able to map predictions from m1
rl.E.y_lin <- raster(,nrow=100,ncols=100,ext=extent(sf.study.area.utm),crs=crs(sf.study.area.utm))
# Make data.frame to be able to make predictions at each pixel (cell of raster)
df.pred <- data.frame(elev = NA,
                      s1 = xyFromCell(rl.E.y_lin,cell=1:length(rl.E.y_lin[]))[,1],
                      s2 = xyFromCell(rl.E.y_lin,cell=1:length(rl.E.y_lin[]))[,2])

# Make spatial predictions at each pixel
df.pred$elev <- predict(m1,df.pred[,2:3])

# View first 6 rows of predictions
head(df.pred) 

# Fill raster file with predictions 
rl.E.y_lin[] <- c(df.pred$elev)

rl.E.y_lin <- mask(rl.E.y_lin,sf.study.area.utm)

# Estimate coordinates and amount of maximum elevation
xyFromCell(rl.E.y_lin,cell=which.max(rl.E.y_lin[]))

rl.E.y_lin[which.max(rl.E.y_lin[])]

# Plot estimate coordinates of maximum elevation
{plot(rl.E.y_lin, main = "Linear with iid errors") # Plot map of predictions
plot(sf.study.area.utm,add=TRUE)
points(xyFromCell(rl.E.y_lin,cell=which.max(rl.E.y_lin[])),col="purple",pch="*",cex=3)
points(xyFromCell(rl.E.y_lin,cell=which.min(rl.E.y_lin[])),col="blue",pch="*",cex=3)}

```

### 2. GAM

```{r message=FALSE, warning=FALSE}
# Try low-rank Gaussian process (i.e., modern kriging model)
library(mgcv)
m1 <- gam(elev~s(s1,s2,bs="gp"),data=df.elev)

# Make raster of study area to be able to map predictions from m1
rl.E.y_gam <- raster(,nrow=100,ncols=100,ext=extent(sf.study.area.utm),crs=crs(sf.study.area.utm))
# Make data.frame to be able to make predictions at each pixel (cell of raster)
df.pred <- data.frame(elev = NA,
                      s1 = xyFromCell(rl.E.y_gam,cell=1:length(rl.E.y_gam[]))[,1],
                      s2 = xyFromCell(rl.E.y_gam,cell=1:length(rl.E.y_gam[]))[,2])

# Make spatial predictions at each pixel
df.pred$elev <- predict(m1,df.pred[,2:3])

# View first 6 rows of predictions
head(df.pred) 

# Fill raster file with predictions 
rl.E.y_gam[] <- c(df.pred$elev)

rl.E.y_gam <- mask(rl.E.y_gam,sf.study.area.utm)

# Estimate coordinates and amount of maximum elevation
xyFromCell(rl.E.y_gam,cell=which.max(rl.E.y_gam[]))

rl.E.y_gam[which.max(rl.E.y_gam[])]

# Plot estimate coordinates of maximum elevation
{plot(rl.E.y_gam, main = "Generalized additive model") # Plot map of predictions
plot(sf.study.area.utm,add=TRUE)
points(xyFromCell(rl.E.y_gam,cell=which.max(rl.E.y_gam[])),col="purple",pch="*",cex=3)
points(xyFromCell(rl.E.y_lin,cell=which.min(rl.E.y_lin[])),col="blue",pch="*",cex=3)}

```

### 3. Regression tree

```{r}
# Try a regression tree instead!
library(rpart)
m1 <- rpart(elev~s1+s2,data=df.elev)

# Make raster of study area to be able to map predictions from m1
rl.E.y_rt <- raster(,nrow=100,ncols=100,ext=extent(sf.study.area.utm),crs=crs(sf.study.area.utm))
# Make data.frame to be able to make predictions at each pixel (cell of raster)
df.pred <- data.frame(elev = NA,
                      s1 = xyFromCell(rl.E.y_rt,cell=1:length(rl.E.y_rt[]))[,1],
                      s2 = xyFromCell(rl.E.y_rt,cell=1:length(rl.E.y_rt[]))[,2])

# Make spatial predictions at each pixel
df.pred$elev <- predict(m1,df.pred[,2:3])

# View first 6 rows of predictions
head(df.pred) 

# Fill raster file with predictions 
rl.E.y_rt[] <- c(df.pred$elev)

rl.E.y_rt <- mask(rl.E.y_rt,sf.study.area.utm)

# Estimate coordinates and amount of maximum elevation
xyFromCell(rl.E.y_rt,cell=which.max(rl.E.y_rt[]))

rl.E.y_rt[which.max(rl.E.y_rt[])]

# Plot estimate coordinates of maximum elevation
{plot(rl.E.y_rt, main = "Regression tree") # Plot map of predictions
plot(sf.study.area.utm,add=TRUE)
points(xyFromCell(rl.E.y_rt,cell=which.max(rl.E.y_rt[])),col="purple",pch="*",cex=3)
points(xyFromCell(rl.E.y_lin,cell=which.min(rl.E.y_lin[])),col="blue",pch="*",cex=3)}

```

### 4. Support vector regression

```{r message=FALSE, warning=FALSE}
# Try support vector regression (or machine)!
library(e1071)
m1 <- svm(elev~s1+s2,data=df.elev)

# Make raster of study area to be able to map predictions from m1
rl.E.y_svr <- raster(,nrow=100,ncols=100,ext=extent(sf.study.area.utm),crs=crs(sf.study.area.utm))
# Make data.frame to be able to make predictions at each pixel (cell of raster)
df.pred <- data.frame(elev = NA,
                      s1 = xyFromCell(rl.E.y_svr,cell=1:length(rl.E.y_svr[]))[,1],
                      s2 = xyFromCell(rl.E.y_svr,cell=1:length(rl.E.y_svr[]))[,2])

# Make spatial predictions at each pixel
df.pred$elev <- predict(m1,df.pred[,2:3])

# View first 6 rows of predictions
head(df.pred) 

# Fill raster file with predictions 
rl.E.y_svr[] <- c(df.pred$elev)

rl.E.y_svr <- mask(rl.E.y_svr,sf.study.area.utm)

# Estimate coordinates and amount of maximum elevation
xyFromCell(rl.E.y_svr,cell=which.max(rl.E.y_svr[]))

rl.E.y_svr[which.max(rl.E.y_svr[])]

# Plot estimate coordinates of maximum elevation
{plot(rl.E.y_svr, main = "Support vector regression") # Plot map of predictions
plot(sf.study.area.utm,add=TRUE)
points(xyFromCell(rl.E.y_svr,cell=which.max(rl.E.y_svr[])),col="purple",pch="*",cex=3)
points(xyFromCell(rl.E.y_lin,cell=which.min(rl.E.y_lin[])),col="blue",pch="*",cex=3)}

```

### 5. Boosted regession tree

```{r message=FALSE, warning=FALSE}
# Try boosted regression tree!
library(gbm)
m1 <- gbm(elev~s1+s2,data=df.elev)

# Make raster of study area to be able to map predictions from m1
rl.E.y_gbm <- raster(,nrow=100,ncols=100,ext=extent(sf.study.area.utm),crs=crs(sf.study.area.utm))
# Make data.frame to be able to make predictions at each pixel (cell of raster)
df.pred <- data.frame(elev = NA,
                      s1 = xyFromCell(rl.E.y_gbm,cell=1:length(rl.E.y_gbm[]))[,1],
                      s2 = xyFromCell(rl.E.y_gbm,cell=1:length(rl.E.y_gbm[]))[,2])

# Make spatial predictions at each pixel
df.pred$elev <- predict(m1,df.pred[,2:3])

# View first 6 rows of predictions
head(df.pred) 

# Fill raster file with predictions 
rl.E.y_gbm[] <- c(df.pred$elev)

rl.E.y_gbm <- mask(rl.E.y_gbm,sf.study.area.utm)

# Estimate coordinates and amount of maximum elevation
xyFromCell(rl.E.y_gbm,cell=which.max(rl.E.y_gbm[]))

rl.E.y_gbm[which.max(rl.E.y_gbm[])]

# Plot estimate coordinates of maximum elevation
{plot(rl.E.y_gbm, main = "Boosted regression tree") # Plot map of predictions
plot(sf.study.area.utm,add=TRUE)
points(xyFromCell(rl.E.y_gbm,cell=which.max(rl.E.y_gbm[])),col="purple",pch="*",cex=3)
points(xyFromCell(rl.E.y_lin,cell=which.min(rl.E.y_lin[])),col="blue",pch="*",cex=3)}

```

